Epistemic Semantics in Guarded String Models

### 2020-06-18 tex/draft.tex
restructuring 1-3

In this discussion, a sequence such as $\bar{q}{p}$ can be viewed as a
valuation of the primitive propositions $q$ and $p$, which is used to
describe world state. The primitives are listed in fixed order, and
left unmarked (indicating true) or marked with the overbar (indicating
false).  Suppose that in the coin example, we had an additional
primitive stative proposition (or atomic test) $t$, interpreted as
tails.  Since a coin is heads or tails but not both, we want to allow
the valuations $h\bar{t}$ and $\bar{h}t$, and disallow $ht$ and
$\bar{h}\bar{t}$.  This is enforced by a {\em state formula},
which is a simply Boolean formula, in this case \exi{effect}{a}.
Where
{\sf{B}} is a set of atomic tests and $\rho$ is a state constraint
over {\sf{B}}, ${\cal{A}}_{\mbox{\sf\tiny{B}}}^{\rho}$ is the set of valuations of
{\sf{B}} that make formula $\rho$ true.

Formulas like the ones in \ex{event1} that describe pre- and
post-conditions are {\em effect formulas}. They are interpreted as
defining relations between valuations, as defined in Figure
\ref{effectfigure}.  The valuations they relate are constrained by the
state formula as well.  For the heads-tails example, let the state
formula and the effect formula for $a_1$ (Aly peeking at heads) be as
specified in \ex{effect}. Then the set of valuations determined by
$\rho$ and the relation on valuations for the event $u$ are
$\sem{\rho}$ and $\sem{\zeta}^{\sem{\rho}}$ as given at the bottom in
\ex{6}.  Given a state formula $\rho$, the the relation characterizing
pre- and post-conditions of an event with effect formula $\zeta$ is
defined to be $\sem{\zeta}^{\sem{\rho}}$.

\begin{example}   \label{effect}
\begin{tabular}[t]{l}
State formula ($\rho$)  $h\bar{t} + \bar{h}t$ \\
Effect formula for $a_1$ ($\zeta$)  $\ev{h}{h}$ \\
\( \sem{\rho} = \{h\bar{t},\bar{h}t\} \)  \\
\( \sem{\zeta}^{\sem{\rho}} = \{\pair{h\bar{t}}{h\bar{t}}\} \)  \\
\end{tabular}
\end{example}

\begin{figure}
\hspace{-2mm}\begin{tabular}{lll}
  state formulas & $\rho,\sigma...$ & $\rho{:}{:}{=} \, \begin{array}[t]{l}\tuple{\mbox{atomic tests}}\,|\\
    0\,|\,1\,| \; \rho{+}\sigma\;| \\ \rho\,{\&}\,\sigma\;|\;\bar{\rho}\end{array}$ \\
  effect formulas & $\zeta,\eta...$ & $\zeta{:}{:}{=} \, \begin{array}[t]{l} \rho\,{:}\,\sigma \,|\, \zeta{+}\eta \\
    \zeta\,{\&}\,\eta\;|\;\bar{\zeta}\end{array}$
\end{tabular}
\vspace{1mm}

\(
\begin{array}{cll}
  \sem{\rho} & = & \mbox{Set of Boolean valuations of} \\
            && \mbox{the atomic tests that satisfy $\rho$.} \\
\sem{\rho\,{:}\,\sigma}^V & = & (\sem{\rho}\cap V)\times(\sem{\sigma}\cap V)\\
\sem{\zeta{+}\eta}^V & = & \sem{\zeta}^V\cup\sem{\eta}^V\\
\sem{\zeta{\&}\eta}^V & = & \sem{\zeta}^V\cap\sem{\eta}^V\\
\sem{\bar{\zeta}}^V & = & V \times V \setminus \sem{\zeta}^V\\
\end{array}  
\)

@@

Kleene Algebra with Tests is an algebraic theory that is defined by
equations and inequalities, which a has model classes including
guarded string models, relational models, finite models, and matrix
models.  This paper focuses on defining a family of concrete guarded
string algebras, the elements of which are sets of guarded strings.
Definitions and notation mostly follow \cite{kozen2001automata}.
Additional syntax and semantics is included to model multi-agent
epistemic semantics.  Guarded strings over a finite alphabet {\sf{P}}
are like ordinary strings, but with truth assignments to a set $T$ of
primitive propositions (primitive tests) alternating with the symbols
from $E$. Following \cite{kozen2001automata}, we call these truth
assignments \emph{atoms}; the set of atoms is notated $\mathcal A$.
In the algebra described by Figure \ref{epikfigure}, $E$ is the set of
events $\{a_1,a_0,b_1,b_0\}$, and in the elevator example, $\{u,d\}$.
In the elevator example, $T$ is $\{p,q\}$, and in the coin example it
is $\{h,t\}$.  We write atoms as the boolean formulae that uniquely
specify the truth assignment. In the elevator example, we have atoms
$\bar q \bar p$, $\bar q p$, $q \bar p$, $q p$, for notational
simplicity we will write these as $\hat0$, $\hat1$, $\hat2$, $\hat3$
respectivelyto indicate the floor each atom represents. In the coin
example, we get $h\bar t$ and $\bar h t$ (for which we use the
shorthand $H$ and $T$), as well as the bogus atoms $\bar h \bar t$ and
$ht$!  To preclude bogosity, the \texttt{constraint} declaration in
Figure~\ref{epikfigure} describes the set of \emph{feasible atoms},
which we parse into a formula $\varphi$ (if there are multiple
\texttt{constraint} declarations, $\varphi$ is the conjunction of the
parsed formulae in each). For a formula $\varphi$, the set of atoms
that satisfy $\varphi$ is $\mathcal A_\varphi$ -- these are the
feasible atoms. In the coin example, the constraint $\varphi$ is
$h\bar t + \bar h t$. Since $\bar h t$ satisfies $\varphi$ and $h \bar
t$ satisfies $\varphi$, they are permitted feasible atoms, and since
$\bar h \bar t$ and $h t$ do not satisfy $\varphi$ they are not
permitted. We omit boolean satisfiability since its definition is
standard.

@@

Fortunately, Epik provides a syntax to specify the input-output
behavior via the \texttt{event} declarations. From these we construct
a function $P$ from agents to relations on atoms that specify the
effects each action can have. If there are multiple event declarations
for the same event, only the latter one is incorporated. To construct
the effect relation $P(e)$ for event $e$, we add a pair $\alpha,
\beta$ to the relation $P(e)$ for every disjunct $\varphi:\varphi'$ such
that $\alpha$ satisfies $\phi$ and $\beta$ satisfies $\varphi'$.  These have the
status of event types, in that they can ``happen'' in different
possible worlds. Here happening corresponds to incrementing the world
with some compatible element of the event type.  Where $e$ is a bare
event symbol, $\llbracket e \rrbracket^{\varphi,P}$ is the
corresponding set of guarded strings, consisting of the bare event
decorated with compatible truth assignments denoted by $P$ that are
admitted by the world-constraints $\varphi$.  See this formally in
Figure~\ref{fig:interp}, and the examples in \ex{events1}.
% u (\bar{q}\bar{p}):(\bar{q}p) + (\bar{q}p):(q\bar{p}) + (q\bar{p}):(qp) 173
% d (\bar{q}p):(\bar{q}\bar{p}) + (q\bar{p}):(\bar{q}p) + (qp):(q\bar{p})

\begin{example} \label{events1}
  \renewcommand{\arraystretch}{1.1}
\(
\begin{array}[t]{l|c}
  e & \llbracket e \rrbracket^{\varphi, P} \\ \hline
  u & \{ \hat 0 ~u~ \hat 1, \hat 1 ~u~ \hat 2,  \hat 2 ~u~ \hat 3\}\\
  d & \{ \hat 1 ~d~ \hat 0, \hat 2~d~\hat 1, \hat 3 ~d~ \hat 2\}\\
  a_1 & \{H ~a_1~ H\}\\
  a_0 & \{T ~a_0~ T\}\\
  b_1 & \{H ~b_1~ H \}\\
  b_0 & \{T ~b_0~ T \}
\end{array}
\)
  \renewcommand{\arraystretch}{1}

@@

@@

Figure~\ref{fig:epik-grammar} depicts the syntax of Epik's formal
term language. As in KAT, $0$ is the null element, $1$ is the unit
element, $+$ indicates disjuction, $;$ indicates event sequencing or
boolean conjunction, $!$ indicates boolean negation, and $*$ indicates
iteration. We add some additional operations to reason about epistemic
alternatives: the term $\Diamond_a~p$ represents epistemic possibility
for agent $a$ in worlds $p$, and complement $\neg p$ represents the
worlds not described by $p$. We can encode $\Box_a~p$, i.e. epistemic
certainty for agent $a$ in worlds $p$, by $\Box_a~p =
(\Diamond_a~p^c)^c$.

@@

  \[\begin{array}{l>{\hspace{-.5em}}c>{\hspace{-0.75em}}l}
  \multicolumn 2 l {\textrm{tests}}& a~\textrm{is an atomic test} \\
  t,s & ::=& a~|~0~|~1~|~t+t~|~t;t~|~!t\\
  \multicolumn 2 l {\textrm{terms}} & e~\textrm{is an atomic event} \\
  p,q & ::=& e~|~0~|~1~|~t~|~p+p~|~p;p~|~p^*~|\neg p~|\Diamond_a~p
  \end{array}\]

  Derived Operators\\
  \mbox{}\qquad \(\Box_a p \triangleq \neg\Diamond_a\neg p \qquad p\&q \triangleq \neg(\neg p + \neg q) \)

@@

We interpret each term in as a set of guarded strings as indicated in
Figure~\ref{fig:interp}. The function $\llbracket p
\rrbracket^{\varphi,P}$ interprets a term $p$ as a set of guarded
strings consistent with state constraints $\varphi$ and effect
relation $P$. A bare event $e$ is interpreted as the set of guarded
strings allowed by $\varphi$ and $P$. The constant $0$ is interpreted
as the empty set.  The constant $1$ is interpreted as the set of all
atoms obeying $\varphi$, i.e. $\{h\bar t, \bar h t\}$ in the coin
example. The operation $p + q$ is interpreted as the union of the
interpretations of $p$ and $q$. The operation $p;q$ interpreted as the
fusion product of the interpretations of $p$ and $q$ raised to sets:
$X \diamond Y = \{ x \diamond y \mid x \in X, y \in Y, x\diamond
y~\textit{defined}\}$.  The operation $p*$ is Kleene star, which is
interpreted to be least upper bound of the iterated fusion product of
the denotation of $p$ with itself. Subsets of $1$ indicated by $t$ are
also elements of $K$, and these form the Boolean algebra of tests,
denoting all atoms admitted by the constraint $\varphi$ and the
boolean test $t$.  We ``overload'' $;$ to further indicate boolean
conjunction, $+$ to further indicate boolean disjunction, and $0,1$ to
indicate the zero and one elements of the boolean algebra. The overbar
operation denotes complement in this Boolean algebra: $\overline t$
denotes all atoms $\alpha$ such that $\alpha \Rightarrow \neg t$.
Other boolean operators (such as implication $\Rightarrow$) are
encoded as syntactic sugar in the standard way. A boolean test $t$
denotes the set of atoms that satisfy $t;\varphi$, that is, they
satisfy both $t$ and $\varphi$.

The unary epistemic alternative operation $\Diamond_a$ for each agent
$a$, is interpreted using using Kripke semantics, as pre-image
relative to a fixed relation $\hat{R_a}$ between guarded strings,
$\Diamond_a~x = \{u \mid \exists v. v \in x \wedge u \mathrel{R_a}
v\}$.  Here $u$ and $v$ are guarded strings, while $x$ is an element
of $K$.

@@

\begin{example}   \label{Rhathat}
\hspace{-2mm}\(
\hat{R} \triangleq
\hspace{-2mm}\begin{array}[t]{l}
\{ \tuple{x,y} \mid \exists u_1 ... \exists u_n \exists v_1 ... \exists v_n. \\
\hspace{6mm} u_1\mathrel{R}v_1 \wedge ... \wedge u_n\mathrel{R}v_n \wedge\\
\hspace{6mm} x \in \llbracket u_1 ; ... ; u_n \rrbracket^{\varphi,P}\hspace{1mm} \wedge \\
\hspace{6mm} y \in \llbracket v_1 ; ... ; v_n \rrbracket^{\varphi,P} \hspace{2mm} \}
\end{array}
\)
\end{example}

\begin{figure}
  \[\begin{array}{l >{\triangleq}c l}
  \llbracket 0 \rrbracket^{\varphi,P} && \{\} \\  
  \llbracket 1 \rrbracket^{\varphi,P} && \{\alpha \mid \alpha \in \mathcal A_\varphi\}\\
  \llbracket e \rrbracket^{\varphi,P}&& \{\alpha~e~\beta \mid
  (\alpha,\beta) \in P(e), \alpha, \beta \Rightarrow \varphi \}
  \\ \llbracket \_ \rrbracket^{\varphi,P}&& \bigcup_{e \in
    \dom(P)}\llbracket e \rrbracket^{\varphi,P}\\
  \llbracket t \rrbracket^{\varphi,P} &&  \{\alpha \mid \alpha \in \mathcal A_{\varphi;t}\}\\
  \llbracket{p+q}\rrbracket^{\varphi,P} && \llbracket{p}\rrbracket^{\varphi,P} \cup \llbracket q \rrbracket^{\varphi,P}\\
  \llbracket{p;q}\rrbracket^{\varphi,P} && \llbracket p \rrbracket^{\varphi,P} \diamond \llbracket{q}\rrbracket^{\varphi,P}\\
  \llbracket p^*\rrbracket^{\varphi,P} && \displaystyle\bigcup_n {\underbrace{\llbracket p \rrbracket^{\varphi,P}\diamond \ldots \diamond \llbracket p \rrbracket^{\varphi,P}}_{n}} \\
  \llbracket \neg p \rrbracket^{\varphi,P} && \llbracket \_^*\rrbracket^{\varphi,P} \setminus \llbracket p \rrbracket^{\varphi,P} \\
  \llbracket \langle a \rangle~p \rrbracket && \{x \mid x~\hat{R_a}~y, y \in \llbracket p \rrbracket\}

@@

The complement operation ${p}^c$ is complement at the level of sets of
guarded strings, defined to every set the universe $\_^*$ except those
denoted by $p$. Note that $\llbracket 0^c \rrbracket^{\varphi,P} =
\llbracket\_^*\rrbracket^{\varphi,P}$. We use the complement to define
universal box modalites as the dual of the diamond modality, $\Box_a x
= (\Diamond_a x^c)^c$.  For instance, Aly is certain that the coin is
heads if and only if she does not consider it possible that it is not
heads.

Summing up, given an Epik program with $n$ agents, we construct state
constraints $\varphi$, effect relations $P$, and a concrete guarded
string model for each term using $\llbracket -
\rrbracket^{\varphi,P}$.
%% $\tuple{K,+,;,*,\bar{},0,1,{}^c,\Diamond_{a_1},...,\Diamond_{a_n}}$.
 $0^c$ is the set of worlds, and it may be countably infinite.
$\Diamond_{a_i}$ is an epistemic modality for the $i$th agent.  Or
referring to the Kripke relations $\hat{R}_i$, the construction
defines a multi-agent Kripke frame
$\tuple{0^c,\hat{R}_1,...,\hat{R}_n}$ (usually a countable one) from
an Epik specification. The frame consists of a set of worlds, and an
epistemic-alternative relation for each agent.  These models are used
as a target for natural-language interpretation in Section 5 and
Section 6, where we obtain semantic values such as \sem{Amy knows that
  it's heads and Bob knows that Amy knows whether it is heads or
  tails, and does not know that it's heads} as elements of
$K$. Concretely the proposositions are sets of guarded strings
(usually countable ones), construed as sets of worlds as they figure
in possible worlds semantics for natural language.

@@

The relation on decorated events needs to be generalized to a relation of worlds.
The principle for this is that an epistemic alternative to a world of the form $we$
is a world of the form $vd$, where $v$ is a world-alternative to $w$, $d$ is an
event-alternative to $e$, and $vd$ is defined (i.e.  the world alternative $v$
satisfies the pre-conditions of the event alternative $d$).  This principle is found
in earlier literature (Moore 198x, Baltag, Moss and Solecki 20xx).  In the
construction in Fst, the definition of world alternatives takes a simple form. Where
$R_a$ is the relation on decorated events for agent $a$, the the corresponding
relation on worlds in is the Kleene closure of $R_a$.  Where $R$ and $S$ are
relations, the concatenation product of $R$ and $S$ is the set of pairs of the form
$\pair{x_1x_2}{y_1y2}$, where $\pair{x_1}{y_1}$ is in relation $R$, and
$\pair{x_2}{y_2}$ is in relation $S$.  The Kleene closure of relation $R$ is
\(\cup_{n \geq 0}R^n\), where $R^n$ is the $n$-times concatenation product of $R$
with itself. This is an operation in the finite state calculus.  Figure
\ref{relclosure} defines the corresponding operation in the guarded string algebra.
The epistemic alternative relation on worlds for an agent is then defined
as the concatenation closure of the event alternative relation for the agent.

@@

We translate the Epik terms into a Haskell algebraic datatype that represents
terms with the same signature as described in Section 2. We then paramaterize
the interpretation function, $\llparenthesis p \rrparenthesis^{\varphi,P}_n$ on an
integer $n$ that describes the maximum length of a string we will produce. The
bounded interpretation into lists of strings is very similar to the unbounded
interpretation into sets of strings, except for the bounds checking. The full
details are shown in Figure~\ref{fig:list-interp}: we use list-comprehension
notation, which is analogous to set-builder notation, except that it produces a
list (potentially with duplicated elements) instead of a set, and so is written
with square brackets $[]$. We use $+$ to indicate list concatenation, and $\sum
l$ to indicate the concatenation of a list of lists $l$. We overload the set
membership symbol $\in$ to also indicate membership in a list. We also lift the
fusion operator $\cdot$ to lists. For a list of guarded strings $l$, we write
$l|_n$ as shorthand for the set $[ g~|~g \in l , |g| \leq n]$.

>>>>

In this discussion, a sequence such as $\bar{q}{p}$ can be viewed as a
valuation of the primitive propositions $q$ and $p$, which is used to
describe world state. The primitives are listed in fixed order, and
left unmarked (indicating true) or marked with the overbar (indicating
false).  Suppose that in the coin example, we had an additional
primitive stative proposition (or atomic test) $t$, interpreted as
tails.  Since a coin is heads or tails but not both, we want to allow
the valuations $h\bar{t}$ and $\bar{h}t$, and disallow $ht$ and
$\bar{h}\bar{t}$.  This is enforced by a {\em state formula}, which is
a simply Boolean formula, in this case \exi{effect}{a}.  Where
{\sf{B}} is a set of atomic tests and $\rho$ is a state constraint
over {\sf{B}}, ${\cal{A}}_{\mbox{\sf\tiny{B}}}^{\rho}$ is the set of
valuations of {\sf{B}} that make formula $\rho$ true. Following
\cite{kozen2001automata}, we call the valuations
\emph{atoms}.

Formulas like the ones in \ex{event1} that describe pre- and
post-conditions are {\em effect formulas}. They are interpreted as
defining relations between valuations, as defined in Figure
\ref{effectfigure}.  The valuations they relate are constrained by the
state formula as well.  For the heads-tails example, let the state
formula and the effect formula for $a_1$ (Aly peeking at heads) be as
specified in \ex{effect}. Then the set of valuations determined by
$\varphi$ and the relation on valuations for the event $u$ are
$\mathcal A_{\mathsf B}^\varphi$ and
$\llbracket\zeta\rrbracket^\varphi$ as given at the bottom in \ex{6}.
Given a state formula $\rho$, the the relation characterizing pre- and
post-conditions of an event with effect formula $\zeta$ is defined to
be $\sem{\zeta}^{\sem{\rho}}$.

\begin{example}   \label{effect}
\begin{tabular}[t]{l}
State formula: $\varphi = h\bar{t} + \bar{h}t$ \\
Effect formula for $a_1$: $\zeta = h \mathbin{:} h$ \\
\( \mathcal{A}^\varphi_{\mathsf B} = \{h\bar{t},\bar{h}t\} \)  \\
\( \sem{\zeta}^\varphi = \{\pair{h\bar{t}}{h\bar{t}}\} \)  \\
\end{tabular}
\end{example}

\begin{figure}
  \[\begin{array}{lcl}
  \multicolumn 3 l {\textrm{state formulas}\hfill (a \in \mathsf{B}) \hfill \mbox{}}\\
  \varphi,\psi &::=& a \mid0 \mid 1 \mid\varphi{+}\psi \mid \varphi~\psi \mid \bar{\varphi} \\
  \multicolumn {3} l {\textrm{effect formulas}}\\
  \zeta,\eta & ::= & \varphi \mathbin{:} \sigma \mid \zeta{+}\eta
    \mid \zeta \mathbin{\&} \eta \mid \bar{\zeta}
  \end{array}\]

\[
\begin{array}{cll}
  \llbracket\psi_1\mathbin{:}\psi_2\rrbracket^\varphi & = &
  \mathcal A^{(\varphi~\psi_1)}_{\mathsf{B}}\times\mathcal A^{(\varphi~\psi_2)}_{\mathsf{B}}\\[.1em]
\llbracket\zeta{+}\eta\rrbracket^\varphi & = & \llbracket\zeta\rrbracket^\varphi\cup\llbracket\eta\rrbracket^\varphi\\
\llbracket\zeta{\&}\eta\rrbracket^\varphi & = & \llbracket\zeta\rrbracket^\varphi\cap\llbracket\eta\rrbracket^\varphi\\
\llbracket\bar{\zeta}\rrbracket^\varphi & = & V \times V \setminus \llbracket\zeta\rrbracket^\varphi\\
\end{array}  
\]

@@

Kleene Algebra with Tests is an algebraic theory that is defined by
equations and inequalities, which a has model classes including
guarded string models, relational models, finite models, and matrix
models.  This paper focuses on defining a family of concrete guarded
string algebras, the elements of which are sets of guarded strings.
Definitions and notation mostly follow \cite{kozen2001automata}.
Additional syntax and semantics is included to model multi-agent
epistemic semantics.  Guarded strings over a finite alphabet {\sf{P}}
are like ordinary strings, but with truth assignments to a set $T$ of
primitive propositions (primitive tests) alternating with the symbols
from $E$.  In the algebra described by Figure \ref{epikfigure}, $E$ is
the set of events $\{a_1,a_0,b_1,b_0\}$, and in the elevator example,
$\{u,d\}$.  In the elevator example, $T$ is $\{p,q\}$, and in the coin
example it is $\{h,t\}$.  We write atoms as the boolean formulae that
uniquely specify the truth assignment. In the elevator example, we
have atoms $\bar q \bar p$, $\bar q p$, $q \bar p$, $q p$, for
notational simplicity we will write these as $\hat0$, $\hat1$,
$\hat2$, $\hat3$ respectivelyto indicate the floor each atom
represents. In the coin example, we get $h\bar t$ and $\bar h t$ (for
which we use the shorthand $H$ and $T$), as well as the bogus atoms
$\bar h \bar t$ and $ht$!  To preclude bogosity, the
\texttt{constraint} declaration in Figure~\ref{epikfigure} describes a
state formula $\varphi$ which denotes the set of \emph{feasible atoms}
$\mathcal A_{\mathsf B}^\varphi$.

@@

Fortunately, Epik provides a syntax to specify the input-output
behavior via the \texttt{event} declarations. From these we construct
a function $\mathcal E$ from agents to relations on atoms that specify
the effects each action can have. Notably for every declaration
$\texttt{event}~e~\zeta$, we have $\mathcal E^\varphi(e) =
\llbracket\zeta\rrbracket^\varphi$. We can lift $\mathcal E^\varphi$
to $\hat{\mathcal E}$ which decorates the events with their permitted
atoms, that is $\hat{\mathcal E}^\varphi(e) = \{\alpha~e~\beta \mid
\alpha, \beta \in \mathcal E^\varphi (e) \}$. See the examples in
\ex{events1}.
% u (\bar{q}\bar{p}):(\bar{q}p) + (\bar{q}p):(q\bar{p}) + (q\bar{p}):(qp) 173
% d (\bar{q}p):(\bar{q}\bar{p}) + (q\bar{p}):(\bar{q}p) + (qp):(q\bar{p})

\begin{example} \label{events1}
  \renewcommand{\arraystretch}{1.1}
\(
\begin{array}[t]{l|c}
  e &  \hat{\mathcal E}^\varphi(e) \\ \hline
  u & \{ \hat 0 ~u~ \hat 1, \hat 1 ~u~ \hat 2,  \hat 2 ~u~ \hat 3\}\\
  d & \{ \hat 1 ~d~ \hat 0, \hat 2~d~\hat 1, \hat 3 ~d~ \hat 2\}\\
  a_1 & \{H ~a_1~ H\}\\
  a_0 & \{T ~a_0~ T\}\\
  b_1 & \{H ~b_1~ H \}\\
  b_0 & \{T ~b_0~ T \}
\end{array}
\)
  \renewcommand{\arraystretch}{1}

@@

\begin{example}   \label{Rhathat}
\hspace{-2mm}\(
\hat{R} \triangleq
\hspace{-2mm}\begin{array}[t]{l}
\{ \tuple{x,y} \mid \exists u_1 ... \exists u_n \exists v_1 ... \exists v_n. \\
\hspace{6mm} u_1\mathrel{R}v_1 \wedge ... \wedge u_n\mathrel{R}v_n \wedge\\
\hspace{6mm} x \in \llbracket u_1 ; ... ; u_n \rrbracket^{\varphi,{\mathcal{E}}}\hspace{1mm} \wedge \\
\hspace{6mm} y \in \llbracket v_1 ; ... ; v_n \rrbracket^{\varphi,{\mathcal{E}}} \hspace{2mm} \}
\end{array}
\)
\end{example}

@@

Figure~\ref{fig:epik-grammar} depicts the syntax of Epik's formal term
language. As in KAT, $0$ is the null element, $1$ is the unit element,
$p + q$ indicates disjuction, $p;q$ (or simply $pq$) indicates
sequence, the boolean operators for tests $\varphi$ are the standard
ones (as in Figure~\ref{fig:effect-syntax}), and ${p}^*$ indicates
iteration of $p$. We add some additional operations to reason about epistemic
alternatives: the term $\Diamond_a~p$ represents epistemic possibility
for agent $a$ in worlds $p$, and complement $\neg p$ represents the
worlds not described by $p$. We can encode $\Box_a~p$, i.e. epistemic
certainty for agent $a$ in worlds $p$, by $\Box_a~p =
(\Diamond_a~p^c)^c$.

@@

  \[\begin{array}{l>{\hspace{-.5em}}c>{\hspace{-0.75em}}l}
  \multicolumn 2 l {\textrm{terms}} & e \in \mathsf{P} \\
  p,q & ::=& e~|~0~|~1~|~\phi~|~p+p~|~p;p~|~p^*~|\neg p~|\Diamond_a~p
  \end{array}\]

  Derived Operators\\[.1em]
  \(\Box_a p \triangleq \neg\Diamond_a\neg p\) \hfill \(p\&q \triangleq \neg(\neg p + \neg q)\) \hfill \(\_ \triangleq \sum\mathsf P\)

@@

We interpret each term in as a set of guarded strings as indicated in
Figure~\ref{fig:interp}. The function $\llbracket p
\rrbracket^{\varphi,{\mathcal{E}}}$ interprets a term $p$ as a set of
guarded strings consistent with state constraints $\varphi$ and effect
relations ${\mathcal{E}^\varphi}$. A bare event $e$ is interpreted as
the set $\hat{\mathcal E}^\varphi(e)$. The constant $0$ is interpreted
as the empty set.  The constant $1$ is interpreted as the set
$\mathcal A^\varphi_{\mathsf B}$, i.e. $\{h\bar t, \bar h t\}$ in the
coin example. The operation $p + q$ is interpreted as the union of the
interpretations of $p$ and $q$. The operation $p;q$ interpreted as the
fusion product of the interpretations of $p$ and $q$ raised to sets:
$X \diamond Y = \{ x \diamond y \mid x \in X, y \in Y, x\diamond
y~\textit{defined}\}$.  The operation $p*$ is Kleene star, which is
interpreted to be least upper bound of the iterated fusion product of
the denotation of $p$ with itself. Subsets of $1$ indicated by
propositions $\psi$ are also elements of $K$, and these form the
Boolean algebra of tests, denoting all atoms admitted by the
constraint $\varphi$ and the boolean test $\psi$.

The unary epistemic alternative operation $\Diamond_a$ for each agent
$a$, is interpreted using using Kripke semantics, as pre-image
relative to a fixed relation $\hat{R_a}$ between guarded strings,
$\Diamond_a~p = \{u \mid \exists v. v \in \llbracket
p\rrbracket^{\phi, \mathcal E} \wedge u \mathrel{\hat{R_a}} v\}$.  Here $u$
and $v$ are guarded strings, while $x$ is an element of $K$.

@@

\begin{figure}
  \[\begin{array}{l >{\triangleq}c l}
  \llbracket 0 \rrbracket^{\varphi,{\mathcal{E}}} && \{\} \\  
  \llbracket 1 \rrbracket^{\varphi,{\mathcal{E}}} && \mathcal A^\varphi_{\mathsf B}\\
  \llbracket e \rrbracket^{\varphi,{\mathcal{E}}}&& \hat{\mathcal E}^\varphi(e) \\
  \llbracket \psi \rrbracket^{\varphi,{\mathcal{E}}} &&  \mathcal A^{\varphi~\psi}\\
  \llbracket{p+q}\rrbracket^{\varphi,{\mathcal{E}}} && \llbracket{p}\rrbracket^{\varphi,{\mathcal{E}}} \cup \llbracket q \rrbracket^{\varphi,{\mathcal{E}}}\\
  \llbracket{p;q}\rrbracket^{\varphi,{\mathcal{E}}} && \llbracket p \rrbracket^{\varphi,{\mathcal{E}}} \diamond \llbracket{q}\rrbracket^{\varphi,{\mathcal{E}}}\\
  \llbracket p^*\rrbracket^{\varphi,{\mathcal{E}}} && \displaystyle\bigcup_n {\underbrace{\llbracket p \rrbracket^{\varphi,{\mathcal{E}}}\diamond \ldots \diamond \llbracket p \rrbracket^{\varphi,{\mathcal{E}}}}_{n}} \\
  \llbracket \neg p \rrbracket^{\varphi,{\mathcal{E}}} && \llbracket \_^*\rrbracket^{\varphi,{\mathcal{E}}} \setminus \llbracket p \rrbracket^{\varphi,{\mathcal{E}}} \\
  \llbracket \Diamond_a~p \rrbracket && \{x \mid x \mathbin{\hat{R_a}} y, y \in \llbracket p \rrbracket^{\varphi,\mathcal E}\}

@@

The complement operation ${p}^c$ is complement at the level of sets of
guarded strings, defined to every set the universe $\_^*$ except those
denoted by $p$. Note that $\llbracket 0^c
\rrbracket^{\varphi,{\mathcal{E}}} =
\llbracket\_^*\rrbracket^{\varphi,{\mathcal{E}}}$.

Summing up, given an Epik program with $n$ agents, we construct state
constraints $\varphi$, effect relations ${\mathcal{E}}$, and a concrete guarded
string model for each term using $\llbracket -
\rrbracket^{\varphi,{\mathcal{E}}}$.
%% $\tuple{K,+,;,*,\bar{},0,1,{}^c,\Diamond_{a_1},...,\Diamond_{a_n}}$.
 $0^c$ is the set of worlds, and it may be countably infinite.
$\Diamond_{a_i}$ is an epistemic modality for the $i$th agent.  Or
referring to the Kripke relations $\hat{R}_i$, the construction
defines a multi-agent Kripke frame
$\tuple{0^c,\hat{R}_1,...,\hat{R}_n}$ (usually a countable one) from
an Epik specification. The frame consists of a set of worlds, and an
epistemic-alternative relation for each agent.  These models are used
as a target for natural-language interpretation in Section 5 and
Section 6, where we obtain semantic values such as \sem{Amy knows that
  it's heads and Bob knows that Amy knows whether it is heads or
  tails, and does not know that it's heads} as elements of
$K$. Concretely the propositions are sets of guarded strings (usually
countable ones), construed as sets of worlds as they figure in
possible worlds semantics for natural language.

@@

The relation on decorated events needs to be generalized to a relation
of worlds.  The principle for this is that an epistemic alternative to
a world of the form $we$ is a world of the form $vd$, where $v$ is a
world-alternative to $w$, $d$ is an event-alternative to $e$, and $vd$
is defined (i.e.  the world alternative $v$ satisfies the
pre-conditions of the event alternative $d$).  This principle is found
in earlier literature (Moore 198x, Baltag, Moss and Solecki 20xx).  In
the construction in Fst, the definition of world alternatives takes a
simple form. Where $R_a$ is the relation on decorated events for agent
$a$, the the corresponding relation on worlds in is the Kleene closure
of $R_a$.  Where $R$ and $S$ are relations, the concatenation product
of $R$ and $S$ is the set of pairs of the form $\pair{x_1x_2}{y_1y2}$,
where $\pair{x_1}{y_1}$ is in relation $R$, and $\pair{x_2}{y_2}$ is
in relation $S$.  The Kleene closure of relation $R$ is \(\cup_{n \geq
  0}R^n\), where $R^n$ is the $n$-times concatenation product of $R$
with itself (the $0$-times concatenation product is $\llbracket 1
\rrbracket^{\varphi,\mathcal E}$). This is an operation in the finite
state calculus.  Figure \ref{relclosure} defines the corresponding
operation in the guarded string algebra.  The epistemic alternative
relation on worlds for an agent is then defined as the concatenation
closure of the event alternative relation for the agent.

@@

We translate the Epik terms into a Haskell algebraic datatype that represents
terms with the same signature as described in Section 2. We then paramaterize
the interpretation function, $\llparenthesis p \rrparenthesis^{\varphi,{\mathcal{E}}}_n$ on an
integer $n$ that describes the maximum length of a string we will produce. The
bounded interpretation into lists of strings is very similar to the unbounded
interpretation into sets of strings, except for the bounds checking. The full
details are shown in Figure~\ref{fig:list-interp}: we use list-comprehension
notation, which is analogous to set-builder notation, except that it produces a
list (potentially with duplicated elements) instead of a set, and so is written
with square brackets $[]$. We use $+$ to indicate list concatenation, and $\sum
l$ to indicate the concatenation of a list of lists $l$. We overload the set
membership symbol $\in$ to also indicate membership in a list. We also lift the
fusion operator $\cdot$ to lists. For a list of guarded strings $l$, we write
$l|_n$ as shorthand for the set $[ g~|~g \in l , |g| \leq n]$.
